---
title: "v4.Clustering_projection"
author: "Genevieve L. Stein-O'Brien"
date: "`r Sys.Date()`"
output: BiocStyle::pdf_document
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\usepackage[utf8]{inputenc} 
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
set.seed(1234)
```

## The base projectoR function

The base projectoR function is executed as follows:
```{r base.projectoR, eval=FALSE}
library(projectoR)
projectoR(data = NA, AnnotionObj = NA, IDcol = "GeneSymbol",
    Patterns = NA, NP = NA, full = FALSE)
```

\par \noindent \textbf{\underline{Input Arguments}} The inputs that must be set each time are only the data and patterns, with all other inputs having default values.  However, inconguities between gene names--rownames of the Patterns object and either rownames of the data object or the values in the "IDcol" of a corresponding annotation object for the data--will throw errors and, subsequently, should be checked before running.  

The arguments are as follows:
\begin{description}
\item[data]{a dataset to be projected into the pattern space}
\item[AnnotionObj]{an annotion object for data. If NA the rownames of data will be used.}
\item[IDcol]{the column of AnnotionData object corresponding to identifiers matching the type used for GeneWeights}
\item[Patterns]{a matrix of continous values with unique rownames to be projected}
\item[NP]{vector of integers indicating which columns of Patterns object to use. The default of NP = NA will use entire matrix.}
\item[full]{logical indicating whether to return the full model solution. By default only the new pattern object is returned.}
\end{description}
